\title{Change Semantics for Semi\naive{} Datafun}
\author{\scshape Michael Arntzenius}
\date{\today}

\maketitle

\begin{abstract}
  We generalize semi\naive{} evaluation from Datalog to Datafun and prove it
  correct.
\end{abstract}

\section{Strategy}

We first generalize Datafun's \Poset{} semantics to any category with
appropriate structures (\cref{sec:datafun-models}). We construct such a
category, \CP{}, whose objects are posets equipped with a notion of increasing
changes, and whose morphisms are monotone maps possessing \emph{derivatives}
relating input and output changes (\cref{sec:changeposets}). We show this
``incremental'' semantics agrees with Datafun's standard semantics
(\cref{sec:incremental}). We give a static transformation on Datafun that
reinterprets the discreteness comonad to compute derivatives, and uses these to
speed up fixed point computations. Finally, we prove this semi\naive{}
evaluation transform correct.

% TODO: more section links once I finish them.


\section{Notation and conventions}

\begin{enumerate}
\item Adjacency $f\<x$ denotes the function $f$ applied to an argument $x$.

\item I write $f \then g$ for the composition of morphisms $f : A \to B$ and $g
  : B \to C$. This is the reverse of the usual composition operator, $g \circ
  f$.

\item I write $\subst{e}{x \substo f}$ to mean ``$e$ with all free occurrences of $x$
  replaced by $f$ ($\alpha$-varying as necessary to avoid capture)''.

  \newcommand\iexpr{\phi}
\item I write $(\iexpr_i)_i$ for the tuple or sequence $(\iexpr_1,
  ..., \iexpr_n)$, leaving $n$ unspecified. For example,

  \nopagebreak[2]
  \[ (2^ix + y_i)_i \quad\text{means}\quad
  (2^1x + y_1,~ 2^2x + y_2,~ ...,~ 2^nx + y_n) \]

  \noindent
  I also use this notation with other braces, eg. $\fork{\iexpr_i}_i =
  \fork{\iexpr_0, ..., \iexpr_n}$.

\item $\expO A B$ denotes the exponential $B^A$ in a cartesian closed category.

\item Given a family of morphisms $f_i : A \to B_i$ in a category with finite
  products, their \emph{``fork''} $\fork{f_i}_i : A \to \textstyle\prod_{i} B_i$
  uniquely satisfies $\fork{f_i}_i \then \pi_i = f_i$. Dually, given $f_i : A_i
  \to B$ in a category with finite sums, their \emph{``split''} $\krof{f_i}_i :
  \sum_i A_i \to B$ uniquely satisfies $\injc_i \then \krof{f_i}_i = f_i$.
  %
  So in particular, $\termI : A \to \termO$ is the unique map into the terminal
  object and $\initE : \initO \to B$ the unique map out of the initial object.

\item The comonad $\iso : \Poset \to \Poset$ takes a poset to the
  \emph{discrete} poset on the same elements, defined:

  \begin{eqnarray*}
    x \in \isof A &\iff& x \in A\\
    x \le y : \isof A &\iff& x = y\\
    \isof(f : A \to B) &=& f : \isof A \to \isof B
  \end{eqnarray*}
\end{enumerate}


\input{seminaive/fig1-core-datafun.tex}
\input{seminaive/fig2-surface-syntax.tex}
\input{seminaive/fig3-desugaring.tex}
\input{seminaive/fig4-desugaring-higher-order-semilattice}
\input{seminaive/fig5-desugaring-pattern-matching}


\section{Surface syntax and its desugaring}

We present Datafun in two layers: a simpler \emph{core}
(\cref{fig:core-datafun}) and a more liberal \emph{surface}
(\cref{fig:surface-syntax}) whose extensions are desugared into core terms
(\crefrange{fig:desugaring}{fig:desugaring-pattern-matching}).
%
The simpler core terms are easier to \emph{analyse}, so our semantics and static
transformations consume core terms; conversely, surface terms are easier to
\emph{construct}, so we use them in example programs and in the outputs of our
static transformations.

The specific surface features we add are:
\begin{enumerate}
\item Let-bindings.

\item Booleans, ordered $\efalse < \etrue$. Since sum types are ordered
  disjointly, booleans are instead desugared into sets of empty tuples, with
  $\efalse = \emptyset$ and $\etrue = \eset{\etuple{}}$. \todo{TODO: explain
    $(\eisEmpty e)$}.

\item Pattern matching on tuples and boxes. \todo{TODO: explain $\prim{split}$.}

%% \item In the core language, semilattice operations ($\bot$, $e \vee f$, and
%%   \kwname{for}-loops) are permitted only at \emph{first-order} semilattice types
%%   $\eqt L$. This limitation is lifted in the surface language using
%%   $\eta$-expansion (\cref{fig:desugaring-higher-order-semilattice}).

\item In the core language, we express fixed point computations using what
  amounts to a higher-order operator $\prim{fix} : \isof{(\fixt L \to \fixt L)}
  \to \fixt L$. In the surface language there is a more convenient binding form,
  $\efixisd x e$.

\end{enumerate}


\section{Datafun models}
\label{sec:datafun-models}

The \emph{``standard''} model of Datafun interprets types as posets and terms as
monotone maps. Generalizing this, we can interpret Datafun terms into any
bicartesian closed category \catC{} equipped with the following structures:

\begin{enumerate}
\item A \emph{``discreteness''} comonad $\tuple{\disco{}, \varepsilon, \delta}$
  equipped with distributive morphisms:

  \nopagebreak[2]
  \begin{align*}
    \discox &: \prod_i \discof{A_i} \to \disco \prod_i A_i
    &
    \discosum &: \disco \sum_i A_i \to \sum_i \discof{A_i}
  \end{align*}

\item A \emph{``finite powerset''} functor \(\pfin : \catC \to \catC\) equipped
  with morphism families:

  \nopagebreak[2]
  \begin{align*}
    \morph{singleton} &: \isof A \to \pfinof{A} &
    \morph{isEmpty} &: \discof{\pfinof{\termO}} \to \termO + \termO
  \end{align*}

  %% \noindent
  %% and, moreover, for any objects $\G,A$, semilattice object $L$, and morphism
  %% $f : \G \x \isof A \to L$, a ``collecting'' morphism \( \pcollect{f} : \G \x
  %% \pfinof A \to L \).

\item \emph{Equality}, \emph{semilattice}, and \emph{fixed point objects}, which
  must interpret first-order, semilattice, and fixed point types respectively.
  Letting $\eqt A$ be an equality object, $L$ be a semilattice object, and
  $\fixt L$ be a fixed point object, these must have morphisms:

  \nopagebreak[2]
  \begin{align*}
    \morph{eq} &: \isof{\eqt A} \x \isof{\eqt A} \to \pfinof\termO
    &
    \morph{join}_n &: L^n \to L
    &
    \morph{fix} &: \isof{(\expO{\fixt L}{\fixt L})} \to \fixt L
  \end{align*}

\item To interpret \kwname{for}-loops, for any objects $\G, A$ and any
  semilattice object $L$ we require a family of morphisms $\pcollect{f}$:

  \nopagebreak[2]
  \[
    \infer{f : \G \x \isof A \to L}
          {\pcollect{f} : \G \x \pfinof{A} \to L}
  \]

%% , Given objects $\G, A$, a semilattice equality type $\eqt L$, and
%%   a morphism $f : \G \x \isof{A} \to \den{\eqt L}$, we require a morphism
%%   $\pcollect{f} : \G \x \pfinof{A} \to \den{\eqt L}$. Or, abusing inference rule
%%   notation:
%%   %
%%   \[
%%     \infer{f : \G \x \isof A \to \den{\eqt L}}
%%           {\pcollect{f} : \G \x \pfinof{A} \to \den{\eqt L}}
%%   \]

\end{enumerate}

\noindent
\Cref{fig:semantics} shows how to interpret Datafun into any Datafun
model. The standard semantics is an instance of this, letting \catC{} be
\Poset{}, \iso{} be \iso{}, and \pfin{} be the finite powerset functor.
\todo{TODO: explain in more detail?}

Since we have not imposed any laws, there is no useful general equational theory
of Datafun models. Instead, our approach is to relate the models we construct to
the standard model and use \emph{its} equational theory.


\input{seminaive/fig6-semantics}


\section{The category \CP}
\label{sec:changeposets}

Objects $A$ of \CP{} are tuples $\tuple{\vals A, \chgs A, \updfn_A,
  \composefn_A}$, where:

\begin{itemize}
\item $\vals A$ is a poset of values.

\item $\chgs A$ is a poset of changes.

\item $\updfn_A : \vals A \x \chgs A \pto \vals A$ is a \emph{partial} map
  taking a value and a change to an updated value. When $x \upd \dx$ is defined,
  we say \dx\ is a \emph{valid} change to $x$. It is often convenient to notate
  $x \upd \dx = y$ relationally, as $\valid\dx x y$ (``\dx\ sends $x$ to $y$''),
  or more briefly as $\vld\dx x y$.

\item $\composefn_A : \chgs A \x \chgs A \to \chgs A : \Poset$ is a monotone
  change composition operator.
\end{itemize}

\noindent
These must satisfy the following conditions:

\nopagebreak[2]
\begin{align*}
  x \le y &\impliedby \ex{\dx} \vld{\dx} x y
  & \text{sound for increases}\\
  x \le y &\implies \ex{\dx} \vld{\dx} x y
  & \text{complete for increases}\\
  \vld\dx x y \wedge \vld\dy y z &\implies \vld{\dx\compose\dy} x z
  & \text{correctness of composition}
\end{align*}

\noindent Completeness for increases is equivalent (via choice) to having an
operator $x \changeto y : \chgs A$, defined for $x \le y : \vals A$, such that
$\valid{x \changeto y} x y$.\footnote{Our $x \changeto y$ is effectively the
  incremental $\fn$-calculus' $y \ominus x$, but defined only when $x \le y$.}
In particular, every $x$ has a \emph{zero change}, $\zero_x = x \changeto x$.
However, changes (including zero changes) are not necessarily unique: for a
given $x,y$ there may be many $\dx$ such that $\valid\dx x y$.

Morphisms $f : \CP(A, B)$ are differentiable maps $f : \Poset(\vals A, \vals
B)$. A map is differentiable if it has a derivative $\deriv f : \vals A \to
\Poset(\chgs A, \chgs B)$ such that:

\nopagebreak[2]
\[ \vld{\dx} x y \implies \vld{\deriv f\<x\<\dx}{f\<x}{f\<y}\]

\noindent Or, equivalently (when $x \upd \dx$ is defined):

\nopagebreak[2]
\[ f\<(x \upd \dx) = f\<x \upd \deriv f\<x\<\dx \]

\noindent
The same morphism may have many different derivatives; all that matters is that
one exist.

Identity and composition morphisms in \CP{} are inherited from \Poset{}, so we
need only exhibit their derivatives. These follow a variant of the chain rule:

\nopagebreak[2]
\begin{align*}
  \deriv\id \<x\<\dx &= \dx
  & \deriv{(g \circ f)} \<x\<\dx &= \deriv g \<(f\<x) \<(\deriv f\<x\<\dx)
\end{align*}

\noindent
It's easy to see these functions are monotone in \dx\ as required.


\subsection{The forgetful functor \valfn}
\label{sec:CP-vals}

The functor $\valfn : \CP \to \Poset$ takes objects $A$ to $\vals A$ and
morphisms $f$ to themselves. This can be seen as a process of ``forgetting''. On
objects, it forgets a poset's associated change-structure; on morphisms, it
forgets their differentiability.
%
Our ``first-order'' constructions in \CP{} --- products, sums, the finite
powerset functor --- commute with \valfn, in a sense \emph{inheriting} from
those in \Poset. For example, products have $\vals{(A \x B)} = \vals A \x \vals
B$ and $\vals{\fork{f,g}} = \fork{\vals f, \vals g}$.
%
Where this applies, we omit for brevity this inherited structure.
%
However, exponentials will prove the necessary exception to this rule: since all
maps in \CP{} are differentiable, so are the elements of $\vals{(\expO A B)}$.

%% The structures we will build in \CP{} will
%% usually \emph{refine} those in \Poset, in the sense of commuting with \valfn.
%% For example, for cartesian products, $\vals{(A \x B)} = \vals A \x \vals B$ and
%% $\vals{\fork{f,g}} = \fork{\vals f, \vals g}$.
%% %
%% For brevity, we omit the ``value components'' of our constructions when they are
%% inherited in this way from \Poset.


%% \subsection{Products}
%% \label{sec:CP-products}

%% \CP\ has all finite products. $\valfn$ and $\chgfn$ distribute across them:

%% \nopagebreak[2]
%% \begin{align*}
%%   \vals{\prod_{i \in I} A_i} &= \prod_{i \in I} \vals A_i &
%%   \chgs{\prod_{i \in I} A_i} &= \prod_{i \in I} \chgs A_i
%% \end{align*}

%% \noindent
%% Updates are pointwise:

%% \nopagebreak[2]
%% \[ (x_i)_i \upd (\dx_i)_i = (x_i \upd \dx_i)_i
%% \quad\text{when the RHS is defined}\]

%% \noindent Soundness and completeness follow pointwise from soundness and
%% completeness of $A_i$; in particular, $(x_i)_i \changeto (y_i)_i = (x_i
%% \changeto y_i)_i$. Composition is also pointwise:

%% \[ (\dx_i)_i \compose (\dy_i)_i = (\dx_i \compose \dy_i)_i \]

%% It's easy to see

%% \noindent \todo{TODO: soundness, completeness, composition, correctness of composition}

%% The projection and fork maps are inherited along \valfn, with their derivatives
%% being:

%% \nopagebreak[2]
%% \begin{align*}
%%   \deriv\pi_i\<x\<\dx &= \pi_i\<\dx &
%%   \deriv{\fork{f_i}_i} &= \fork{\deriv f_i}_i
%% \end{align*}

%% \noindent Correctness of $\deriv\pi_i$ is straightforward; if
%% $\valid{(\dx_1,\dx_2)}{(x_1,x_2)}{(y_1,y_2)}$ then by definition
%% $\valid{\dx_i}{x_i}{y_i}$. As for correctness of $\deriv{\fork{f_i}_i}$:

%% \nopagebreak[2]
%% \begin{equation*}
%%   \fork{f_i}_i \<(x \upd \dx)
%%   = (f_i(x \upd \dx))_i
%%   = (f_i\<x \upd \deriv f_i \<x \<\dx)_i
%%   = \fork{f_i}_i \<x \upd \fork{\deriv f_i}_i \<x \<\dx
%% \end{equation*}

%% \noindent
%% Finally, the universal property of $\fork{f_i}_i$ follows from its universal
%% property in \Poset\ --- requiring differentiability merely makes the property
%% weaker.


%% \subsection{Sums}
%% \label{sec:CP-sums}

%% \CP\ has all finite sums. \valfn\ and \chgfn\ distribute across them:

%% \nopagebreak[2]
%% \begin{align*}
%%   \vals{\sum_{i \in I} A_i} &= \sum_{i \in I} \vals A_i &
%%   \chgs{\sum_{i \in I} A_i} &= \sum_{i \in I} \chgs A_i
%% \end{align*}

%% \noindent
%% Updates are pointwise but require that the tags match:

%% \nopagebreak[2]
%% \[ \inj i x \upd \inj j \dx =
%% \begin{cases}
%%   \inj i (x \upd \dx) & \text{if}~i=j\\
%%   \text{undefined} & \text{otherwise}
%% \end{cases}
%% \]


\subsection{Cartesian structure}
\label{sec:CP-cartesian}
\input{seminaive/fig-changeposet-cartesian}

\CP\ has all finite products and sums (\cref{fig:CP-cartesian}). Their
value-level structure (including $\pi_i$, $\injc_i$, $\fork{f_i}_i$, and
$\krof{f_i}_i$) is inherited from \Poset; \valfn\ and \chgfn\ distribute across
both products and sums. Soundness and completeness follow directly from the
pointwise ordering of products and sums in \Poset{} and soundness \&
completeness for each factor/summand $A_i$. In particular, \changetofn\ can be
defined by:

\nopagebreak[2]
\begin{align*}
  (x_i)_i \changeto (y_i)_i &= (x_i \changeto y_i)_i &
  \inj i x \changeto \inj i y &= \inj i (x \changeto y)
\end{align*}

\noindent The universal properties of $\fork{f_i}_i$ and $\krof{f_i}_i$ follow
from the corresponding universal properties in \Poset\ --- requiring
differentiability merely weakens them.
%
As for correctness of composition and the derivatives of $\pi_i$/$\injc_i$ and
$\fork{f_i}_i$/$\krof{f_i}_i$, we will show these in turn for products and sums.

\subsubsection{Products}

Correctness of change composition is straightforward: if
%
\[ \vld{(\dx_i)_i}{(x_i)_i}{(y_i)_i}
\quad\text{and}\quad
\vld{(\dy_i)_i}{(y_i)_i}{(z_i)_i} \]
%
then for each $i$ we have $\valid{\dx_i \compose \dy_i}{x_i}{z_i}$ and therefore
%
\[\valid{(\dx_i)_i \compose (\dy_i)_i = (\dx_i \compose
  \dy_i)_i}{(x_i)_i}{(z_i)_i}\]

\noindent
Likewise for correctness of $\deriv\pi_i$; if $\valid{\dx_i}{x_i}{y_i}$ for each
$i$ then by definition
\[\valid{\pi_i\<(\dx_i)_i = \dx_i}{x_i}{y_i}\]

\noindent As for correctness of $\deriv{\fork{f_i}_i}$:

\nopagebreak[2]
\begin{equation*}
  \fork{f_i}_i \<(x \upd \dx)
  = (f_i(x \upd \dx))_i
  = (f_i\<x \upd \deriv f_i \<x \<\dx)_i
  = \fork{f_i}_i \<x \upd \fork{\deriv f_i}_i \<x \<\dx
\end{equation*}

\subsubsection{Sums}

Change composition for sums is interesting, because it must handle changes
$(\inj i \dx \compose \inj j \dy)$ with disjoint tags $i \ne j$. Recall that
composition is correct if, when $\valid\dx x y$ and $\valid\dy y z$, we have
$\valid{\dx\compose\dy} x z$. But if $i \ne j$, then
%
there \emph{are} no $x,y,z$ such that $\valid{\inj i \dx} x y$ and $\valid{\inj j \dy} y z$.
%
This case is \emph{dead code}; the only guarantee we need is monotonicity with
respect to \dx\ and \dy, so it suffices to simply return the left argument.

Correctness of the other case is straightforward: since $i = j$, from the
definition of $\updfn_{\sum_i A_i}$ we know all the tags must agree and we have:

\nopagebreak[2]
\[ \inj i x \longvalidarrow{\inj i \dx} \inj i y
\longvalidarrow{\inj i \dy} \inj i z \]

\noindent From this it follows that \( {x} \longvalidarrow{\dx} {y}
\longvalidarrow{\dy} {z} \) and therefore $\vld{\dx\compose\dy} x z$ and so
finally

\nopagebreak[2]
\[ \valid{\inj i \dx \compose \inj i \dy = \inj i (\dx \compose \dy)}{\inj i x}{\inj i z} \]

\noindent
Correctness of $\deriv\injc_i$ is immediate: suppose $\vld\dx x x$. Then
\(\vld{\inj i \dx}{\inj i x}{\inj i x}\). As for split, its derivative must obey

\nopagebreak[2]
\begin{equation*}
  \krof{f_i}_i\<(\inj i x) \upd \deriv{\krof{f_i}_i} \<(\inj i x) \<(\inj i \dx)
  = \krof{f_i}_i\<(\inj i x \upd \inj i \dx)
  = f_i \<(x \upd \dx)
\end{equation*}

\noindent It's easy to see that the $i = j$ case of the definition in
\cref{fig:CP-cartesian} satisfies this property. As with $\composefn_{\sum_i
  A_i}$, the $i \ne j$ case is dead code.


%% \subsection{Cartesian structure}
%% \label{sec:CP-cartesian}

%% \todo{TODO: Neel suggests splitting this into products \& sums. It's easier to
%%   understand the quirks of sums on their own rather than while also trying to
%%   understand products.}

%% \CP\ has all finite products and sums. The value-level structure is inherited
%% from \Poset\ along \valfn. The change-poset structure is also inherited from
%% \Poset, while the projection and injection's derivatives operate pointwise on
%% changes:

%% \nopagebreak[2]
%% \begin{align*}
%%   \chgs{\prod_{i \in I} A_i} &= \prod_{i \in I} \chgs A_i &
%%   \chgs{\sum_{i \in I} A_i} &= \sum_{i \in I} \chgs A_i
%%   %\label{eqn:delta-distributes-over-sums-and-products}
%%   \\
%%   \deriv\pi_i\<x\<\dx &= \pi_i\<\dx & \deriv\injc_i\<x\<\dx &= \inj i \dx
%% \end{align*}

%% \noindent \todo{TODO: derivatives of fork \& split!}

%% \noindent Updates are pointwise, noting that $(\inj i x \upd \inj j \dx)$ is
%% undefined for $i \ne j$:

%% \nopagebreak[2]
%% \begin{align*}
%%   (x_i)_i \upd (\dx_i)_i &= (x_i \upd \dx_i)_i &
%%   \inj i x \upd \inj i \dx &= \inj i (x \upd \dx)
%% \end{align*}

%% \noindent
%% Soundness \& completeness for increases follow directly from the pointwise
%% orderings of products \& coproducts in \Poset{}. In particular, the $\changetofn$
%% operator is given by:

%% \nopagebreak[2]
%% \begin{align*}
%%   (x_i)_i \changeto (y_i)_i &= (x_i \changeto y_i)_i &
%%   \inj i x \changeto \inj i y &= \inj i (x \changeto y)
%% \end{align*}

%% \noindent Finally, change composition is also pointwise --- mostly:

%% \nopagebreak[2]
%% \begin{align*}
%%   (\dx_i)_i \compose (\dy_i)_i &= (\dx_i \compose \dy_i)_i &
%%   \inj i \dx \compose \inj j \dy &=
%%   \begin{cases}
%%     \inj i (\dx \compose \dy) &\text{if}~i=j\\
%%     \inj i \dx &\text{otherwise}
%%   \end{cases}
%% \end{align*}

%% \noindent
%% However, for sums we must also handle changes with distinct tags. Recall that
%% composition is correct if, when $\valid\dx x y$ and $\valid\dy y z$, we have
%% $\valid{\dx\compose\dy} x z$. But if $i \ne j$, then
%% %
%% there \emph{are} no $x,y,z$ such that $\valid{\inj i \dx} x y$ and $\valid{\inj j \dy} y z$.
%% %
%% This case is \emph{dead code}; the only guarantee we need is monotonicity with
%% respect to \dx\ and \dy, so it suffices to simply return the left argument.

%% Correctness of the other cases follows directly from correctness of each $\composefn_{A_i}$.


\subsection{Exponentials}
\label{sec:CP-exponentials}

The values $\vals{(\expO A B)}$ of the exponential object are the
\emph{differentiable} monotone maps $\CP(A, B)$, ordered pointwise. Because we
require differentiability, \valfn\ does \emph{not} commute with exponentials:
$\vals{(\expO A B)} \ne \expO{\vals A}{\vals B}$.
%
Meanwhile, the changes $\chgs{(\expO A B)}$ are monotone maps $\isof{\vals A}
\to \chgs A \to \chgs B$, ordered pointwise. The update relation and change
composition must be very carefully defined, so we examine them at length below.

\subsubsection{Updating functions}

The update relation is:

\nopagebreak[2]
\begin{equation}\label{eqn:function-update}
  \vld{\df}{f}{g} : \expO A B
  \iff \fa{\vld\dx x y} \vld{\df\<x\<\dx} {f\<x} {g\<y}
\end{equation}

\noindent However, recall that $\upd$ must be a partial function: supposing
$\valid\df f g$ and $\valid\df f h$, we must show $g = h$. So consider any $x :
A$. \Cref{eqn:function-update} applied to $\valid{\zero_x} x x$ yields

\nopagebreak[2]
\[ \vld{\df \<x \<\zero_x}{f \<x}{g\<x}
\quad\text{and}\quad \vld{\df \<x \<\zero_x}{f \<x}{h\<x} \]

\noindent
Then by functionality of $\updfn_B$ we have $g\<x = h\<x$, and thus $g = h$.

Although this shows that \cref{eqn:function-update} defines a partial function,
one may wonder what it actually \emph{does} (when it is defined). So let's
rewrite \cref{eqn:function-update} in terms of $\upd_{\expO A B}$:

\nopagebreak[2]
\[ f \upd \df = g \iff \fa{x \upd \dx = y} f\<x \upd \df\<x\<\dx = g\<y \]

\noindent Substituting away $g$ and $y$, this implies that for all valid
\dx, \df:

\nopagebreak[2]
\[
f\<x \upd \df\<x\<\dx = (f \upd \df) \<(x \upd \dx)
\]

\noindent In particular, letting $\dx = \zero_x$ and swapping sides, we get:

\nopagebreak[2]
\[ (f \upd \df) \<x = f\<x \upd \df\<x\<\zero_x \]

\noindent
\todo{TODO: explain why this does not suffice as a \emph{definition} of
  $\updfn_{\expO A B}$ --- in particular, they don't explain when $f \upd \df$
  is defined! It's \emph{not} simply when $x \mapsto f\<x \upd \df\<x\<\dx$ is
  monotone, for example. Can I find a counterexample?}

\subsubsection{Soundness and completeness}

Soundness follows from completeness at $A$ and soundness at $B$: if $\valid\df f
g$ then for any $x$ we have $\valid{\df\<x\<\zero_x}{f\<x}{g\<x}$ (using
completeness of $A$ to find $\zero_x$), and by soundness at $B$ we have $f\<x
\le g\<x$, so $f \le g$.

We'll show completeness by constructing \changetofn\ explicitly:

\nopagebreak[2]
\[ (f \changeto_{\expO A B} g) \<x \<\dx = (f\<x \changeto_B g\<x) \compose_B \deriv g\<x\<\dx\]

\noindent Observe that $f\<x \changeto_B g\<x$ is defined because $f \le g$
implies $f\<x \le g\<x$, and the RHS is monotone in \dx\ by monotonicity of
$\deriv g \<x$ and of $\composefn_B$.
%
To prove completeness, assuming $\valid\dx x y$, we need to show \( \valid{(f
  \changeto g) \<x \<\dx}{f\<x}{g\<y} \). This follows from correctness of
$\composefn_B$, since \( \valid{f\<x \changeto g\<x}{f\<x}{g\<x} \) and
\(\valid{\deriv g \<x \<\dx}{g\<x}{g\<y}\) by definition.

\subsubsection{Function change composition}

Change composition is defined by:

\nopagebreak[2]
\[ (\df \compose_{\expO A B} \dg) \<x \<\dx = \df \<x \<\zero_x \compose_B \dg \<x \<\dx\]

\noindent
This is monotone in \dx\ by monotonicity of $\dg\<x$ and $\composefn_B$. To show
this valid, suppose $\vld\df f g$, $\vld\dg g h$ and $\vld\dx x y$. Then
%
\(\vld{\df\<x\<\zero_x}{f\<x}{g\<x}\)
\text{and}
\(\vld{\dg\<x\<\dx}{g\<x}{h\<y} \).
%
\noindent So by correctness of $\composefn_B$ we have \(\valid{(\df\<x\<\zero_x
  \compose \dg\<x\<\dx)}{f\<x}{h\<y}\) as desired.

\todo{TODO: discuss unavoidable asymmetry in definitions of \(\changetofn_{\expO
    A B}\) and \(\composefn_{\expO A B}\).}

\subsubsection{\eval\ and \fn}

To show that $\expO A B$ is an exponential object, we need evaluation and
currying maps, \eval\ and $\fn f$. Modulo differentiability, these are
inherited from \Poset{}, with the following derivatives:

\begin{align*}
  %% \eval &: (\expO A B) \x A \to B &
  %% \fn (f : \G \x A \to B) &: \G \to \expO A B
  %% \\
  \vals{\eval} \<(f,x) &= f\<x &
  \vals{(\fn f)} \<x &= y \mapsto f\<(x,y)\\
  \deriv{\eval} \<(f,x) \<(\df,\dx) &= \df \<x \<\dx &
  \deriv{(\fn f)} \<x \<\dx &= y\<\dy \mapsto \deriv f \<(x,y) \<(\dx,\dy)
\end{align*}

\noindent
$\deriv{\eval}$ is correct by the definition of $\updfn_{\expO A B}$, for if
$\valid\df f g$ and $\valid\dx x y$ then

\nopagebreak[2]
\[ \eval\<(f,x) = {f\<x} \longvalidarrow{\df\<x\<\dx} {g\<y} = \eval\<(g,y) \]

\noindent
Similarly, \(\deriv{(\fn f)}\) is correct, for suppose $\valid\dx {x_1}{x_2}$
and $\valid\dy{y_1}{y_2}$. Then by differentiability of $f$ we have

\nopagebreak[2]
\[ \fn f \<x_1 \<y_1 =
f\<(x_1,y_1) \longvalidarrow{\deriv f \<(x_1,y_1) \<(\dx,\dy)} f\<(x_2,y_2)
= \fn f \<x_2 \<y_2
\]

\noindent
Finally, $\fn f$ is unique such that $(\fn f \x \id) \then \eval = f$ by the
same argument as in \Poset{} (or indeed \cat{Set}): supposing $\eval \<(\fn f
\<x, y) = f\<(x,y) = \eval\<(g\<x,y)$ for all $x,y$, then we have $\fn f \<x \<
y = g \<x \<y$ and so $\fn f = g$.


\subsection{The pseudo-exponential}

\newcommand\pseudoexp{\rightrightarrows}
%\renewcommand\pseudoexp\rightarrowtriangle
%\renewcommand\pseudoexp{\overset{?}{\Rightarrow}}

You might have expected changes $\chgs{(\expO A B)}$ to be given pointwise, as
maps $\isof{\vals A} \to \chgs B$. Intriguingly, this does yield a valid change
poset, one rather simpler than $\expO A B$, but it does not satisfy the
properties of an exponential. I call this the \emph{pseudo-exponential}, $A
\pseudoexp B$. Pseudo-exponentials are irrelevant to Datafun, but are an
interesting and surprising aspect of \CP, so I describe them here. They are
constructed as follows:

\vspace{-1.2ex}
\begin{mathpar}
  \vals{(A \pseudoexp B)} = \CP(A,B)

  \chgs{(A \pseudoexp B)} = \isof{\vals A} \to \chgs B

  \vld{\df} f g : A \pseudoexp B \iff \fa{x} \vld{\df\<x}{f\<x}{g\<x}

  (\df \compose_{A \pseudoexp B} \dg) \<x = \df\<x \compose_B \dg\<x

  (f \changeto g) \<x = f\<x \changeto g\<x
\end{mathpar}

\noindent
I leave it as an exercise for the reader to verify soundness \& completeness and
correctness of composition. \todo{TODO: explain why they aren't the exponential
  object and aren't isomorphic to $\expO A B$.}

%% Besides $\expO A B$, there is another, simpler, change poset which one might
%% expect to be the exponential, but which is not, which I term the
%% \emph{pseudo-exponential}, written $A \pseudoexp B$.


\subsection{The lifted \iso\ comonad}
\label{sec:CP-iso}

Now we construct a comonad $\iso$ on \CP\ such that $\valfn\iso_\CP =
\iso_\Poset\valfn$. This requirement defines the value-level structure, leaving
only the change poset, update map, and composition. Since $\isof A$ is ordered
discretely, values \emph{cannot} increase. So the simplest definition is:

\nopagebreak[2]
\begin{align*}
  \chgs{\isof A} &= \termO &
  x \upd_{\isof A} \tuple{} &= x &
  \tuple{} \compose_{\isof A} \tuple{} &= \tuple{}
\end{align*}

\noindent
This trivially satisfies soundness, completeness, and correctness of
composition.
%
Functoriality, extraction, duplication, and product- and sum-distribution
inherit from the corresponding properties of $\iso_\Poset$, and their
derivatives are fairly straightforward:

\nopagebreak[2]\vspace{-1ex}
\begin{mathpar}
  \deriv{\iso(f)} \<x \<\tuple{} = \tuple{}

  \deriv\delta_A \<x \<\tuple{} = \tuple{}

  \deriv\varepsilon_A \<x \<\tuple{} = \zero_x
  \\
  \deriv{{\discox}} \<x \<\dx = \tuple{}

  \deriv{{\discosum}} \<(\inj i x) \<\tuple{} = \inj i {\tuple{}}
\end{mathpar}


\subsubsection{The alternative \altiso}

Although we will not use it, there is another refinement of $\iso$ in \CP{},
which I'll call $\altiso$ for clarity, defined like so:

\nopagebreak[2]
\begin{align*}
  \chgs{\altiso A} &= \isof{\chgs A} &
  %% x \upd_{\altiso A} \dx = y &\iff x = y \wedge x \upd_A \dx = y
  \vld{\dx} x x : \altiso A &\iff \vld{\dx} x x : A
  &
  \composefn_{\altiso A} &= \composefn_A
\end{align*}

\noindent
So $\altiso A$ inherits all of $A$'s changes, but ordered discretely and with
\emph{only zero-changes still valid}. Soundness and completeness for increases
and correctness of composition are all implied by the corresponding properties
of $A$. The derivatives of functoriality, duplication, and extraction are
different:

\nopagebreak[2]
\begin{align*}
  \deriv{\altiso(f)} &= \deriv f
  & \deriv\delta_A \<x\<\dx &= \dx
  & \deriv\varepsilon_A \<x\<\dx &= \dx
\end{align*}

\noindent
Observe that we no longer need the non-constructive $\zero_x$ to define comonad
extraction $\varepsilon$! However, for our purposes this will not be an issue.
Meanwhile, product- and sum-distribution become identity maps, because $\prod_i
\altiso A_i = \altiso \prod_i A$ and $\sum_i \altiso A_i = \altiso \sum_i
A_i$ (checking this is left as an exercise for the reader).


\subsection{The lifted \pfin\ functor}
\label{sec:CP-pfin}

We lift the finite powerset functor \pfin\ to \CP\ as follows:

\nopagebreak[2]
\begin{align}
  \vals{\pfinof A} &= \pfinof{\vals A} &
  \chgs{\pfinof A} &= \pfinof{\vals A} &
  (\updfn_{\pfinof A}) = (\composefn_{\pfinof A}) = ({\cup})
  \label{eqn:CP-pfin}
\end{align}

\noindent Since $\updfn$ and \composefn\ are just $\cup$, soundness and
completeness come from a standard property of union (or more generally of
semilattice join):

\nopagebreak[2]
\[ x \le z \iff \ex{y} x \cup y = z \]

\noindent Meanwhile, correctness of composition is simply associativity of
$\cup$:

\nopagebreak[2]
\[ (x \cup \dx = y) \wedge (y \cup \dy = z) \implies (x \cup \dx \cup \dy = z) \]

\noindent
Since the arguments to \morph{singleton} and \morph{isEmpty} cannot change
(having $\iso$ type), their derivatives always return zero-changes:

\nopagebreak[2]
\begin{align*}
  \deriv{\morph{singleton}}_A \<x \<\tuple{} &= \emptyset &
  \deriv{\morph{isEmpty}} \<x \<\tuple{} &= \morph{isEmpty} \< x
\end{align*}



\subsection{Equality objects}
\label{sec:CP-eq}

Every object in \CP{} is an equality object, with \morph{eq} and its derivative
defined by:

\nopagebreak[2]
\begin{align*}
  \morph{eq}\<\tuple{x,y} &= 
  \begin{cases}
    \{\tuple{}\} & \text{if}~ x = y\\
    \emptyset & \text{otherwise}
  \end{cases}
  &
  \deriv{\morph{eq}} \<\pwild \<\pwild &= \emptyset
\end{align*}

\noindent Note that $\deriv{\morph{eq}}$ is correct because $\morph{eq}$'s
arguments cannot change, having $\iso$ type.

Since \emph{every} object is an equality object, every equality type
is trivially interpreted into an equality object. \todo{TODO more explanation
  --- why isn't every type an equality type, then?}


\subsection{Semilattice objects and collecting morphisms}
\label{sec:CP-semilattice}

A semilattice object $L$ in \CP\ is one where $\vals L = \chgs L$ is a
semilattice and $x \upd_L y = x \vee y$. \todo{(TODO: we might want to require
  $\composefn_L = {\vee}$ as well, if we need it.)} This immediately justifies
letting $\deriv{\morph{join}} \<x \<\dx = \morph{join}\<\dx$, because:

\nopagebreak[2]
\begin{align*}
  \morph{join}\<(x \upd \dx) &= \bigvee_i \pi_i \<(x \vee \dx)\\
  &= \left(\bigvee_i \pi_i\<x\right) \vee \left(\bigvee_i \pi_i\<\dx\right)\\
  &= \morph{join}\<x \vee \morph{join}\<\dx\\
  &= \morph{join}\<x \upd \deriv{\morph{join}}\<x\<\dx
\end{align*}

\begin{restatable}{lemma}{lemSemi}
  \label{lem:semi} $\incdens L$ is a semilattice object.
\end{restatable}
\begin{proof}
  By induction on $L$. See \cpageref{proof:semi}.
\end{proof}


\subsection{Collecting morphisms}
\label{sec:CP-collect}

Let $\G, A$ be objects, $L$ be a semilattice object, and $f : \G \x \isof A \to
L$. Our derivative for $\pcollect f : \G \x \pfinof A \to L$ is:

\nopagebreak[2]
\begin{align*}
  \deriv{\pcollect f} \<\tuple{\gamma,s} \<\tuple{\dgamma,\ds}
  &= \bigvee_{x \in \ds} f\<\tuple{\gamma,x} \vee
  \bigvee_{x \in s \cup \ds} \deriv f \<\tuple{\gamma,x} \<\tuple{\dgamma, \zero_x}
\end{align*}

\noindent To establish that this is correct, suppose
$\vld{\dgamma}{\gamma}{\gamma'} : \G$. From this we have:

\nopagebreak[2]
\begin{align*}
  &\phantom{{}={}}\pcollect{f} \<\tuple{\gamma', s \cup \ds}
  \\
  &= \bigvee_{x \in s \cup \ds} f \<\tuple{\gamma', x}
  \\
  &\overset{\footnotemark}{=} \bigvee_{x \in s \cup \ds} \bigl(f \<\tuple{\gamma,x}
  \vee \deriv f \<\tuple{\gamma,x} \<\tuple{\dgamma,\zero_x}\bigr)
  \\
  &= \bigvee_{x \in s} f\<\tuple{\gamma,x}
  \upd \bigvee_{x \in \ds} f\<\tuple{\gamma,x}
  \vee \bigvee_{x \in s \cup \ds} \deriv f \<\tuple{\gamma,x} \<\tuple{\dgamma,\zero_x}
  \\
  &= \pcollect f \<\tuple{\gamma,s}
  \upd \deriv{\pcollect f} \<\tuple{\gamma,s} \<\tuple{\dgamma,\ds}
\end{align*}

\nopagebreak% for footnote to be on same page as mark
\noindent
Which is what we wished to show.\footnotetext{This equality follows from the
  correctness of $\deriv f$ applied to
  \(\vld{\tuple{\dgamma,\zero_x}}{\tuple{\gamma,x}}{\tuple{\gamma',x}}\).}


\subsection{Fixed point objects}
\label{sec:CP-fix}

A fixed point object $\fixt L$ in \CP\ is a semilattice object such that
$\vals{\fixt L}$ satisfies the ascending chain condition. \todo{TODO more
  explanation.} Since the argument to \morph{fix} is boxed and cannot change,
$\deriv{\morph{fix}}$ needs to produce a zero change; and at any semilattice
object, $\bot$ is a zero change, so \( \deriv{\morph{fix}} \<x \<\tuple{} = \bot
\).

\begin{lemma}\label{lem:fixobject}
  \(\incdens{\fixt L}\) is a fixed point object.
\end{lemma}
\begin{proof}
  By induction on $\fixt L$. \todo{TODO}
\end{proof}


\section{Incrementalization}
\label{sec:incremental}

\newcommand\agrees\approx

The \emph{incremental semantics} for Datafun, notated $\incden{-}$, is given by
the Datafun model in \CP{} constructed in \cref{sec:changeposets}, letting
$\disco = \iso_\CP$ and $\pfin = \pfin_\CP$. The obvious question is: how does
this semantics relate to the standard \Poset\ semantics, $\den{-}$? Recalling
\cref{sec:CP-vals}, since most of our constructions inherit along $\valfn$,
restricted to the \Poset-level they are very nearly the same!
%
However, $\vals{\incden{A \to B}} \ne \den{A \to B}$ because of the
differentiability restriction on \CP's exponentials.
%
\todo{TODO: more explanation}

\begin{restatable}[First-order semantic agreement]{theorem}{thmAgreeFirstOrder}
  \label{thm:agree-first-order}
  \( \vals{\incdens{\eqt A}} = \den{\eqt A} \).
\end{restatable}
\begin{proof}
  \todo{TODO, by induction on types}
\end{proof}

\input{seminaive/fig-agrees}

\begin{restatable}[First-order agreement is equality]{theorem}{thmAgreeEqualsEquals}
  \(x \agrees_{\eqt A} y \iff x = y\).
\end{restatable}
\begin{proof}
  \label{thm:agree-equals-equals}
  \todo{TODO, by induction on eqtypes.}
\end{proof}

\begin{restatable}[Semantic agreement]{theorem}{thmAgree}
  If $\J e \G A$
  %and $\g \agrees_\G \sigma$ then $\incdens{e}\<\g \agrees_A \den{e}\<\sigma$.
  then $\incdens{e} \agrees_{\G \vdash A} \den{e}$.
\end{restatable}
\begin{proof}
  \todo{TODO by induction on typing derivations}
\end{proof}

\noindent
The change poset $\D\incdens A$ of a Datafun type $A$ is internally definable
via a type translation given in \cref{fig:PhiDelta}, also named $\D$, despite
the risk of confusion. \todo{TODO: move this \P\ into \cref{sec:seminaive}.}

\input{seminaive/fig7-PhiDelta}

\begin{restatable}{theorem}{thmDeltaDen}
  \label{thm:delta-den}
  \(\D\incdens A = \vals{\incdens{\D A}}\).
\end{restatable}
\nopagebreak[2]
\begin{proof}
  \label{proof:delta-den}
  By induction on types. A starred equality, $\stareq$, marks equalities given by
  an inductive hypothesis; additionally, $\daggereq$ indicates an appeal to
  \cref{thm:agree-first-order}.
  \begin{description}
    \item[Case $\isof A$:] \(\D\incdens{\isof A} = \D{\isof{\incdens A}} =
      \termO_\Poset = \vals\termO_\CP = \vals{\incdens{\tunit}} \).

    \item[Case $\tseteq A$:]
      \(\D\incdens{\,{\tseteq A}\,} = \D\pfinof{\incdens{\eqt A}}
      = \pfinof{\vals{\incdens{\eqt A}}}
      = \vals{\pfinof{\incdens{\eqt A}}}
      = \vals{\incdens{\,{\tseteq A}\,}}
      = \vals{\incdens{\D{\tseteq A}}}
      \).

    \item[Case $A \x B$:]
      \(\D\incden{A \x B} = \D(\incdens A \x \incdens B)
      = \D\incdens A \x \D\incdens B
      \stareq \vals{\incdens{\D A}} \x \vals{\incdens{\D B}}
      = \vals{\incden{\D A \x \D B}}
      = \vals{\incdens{\D(A \x B)}}
      \).

    \item[Case $A + B$:]
      \(\D\incden{A + B} = \D(\incdens A + \incdens B)
      = \D\incdens A + \D\incdens B
      \stareq \vals{\incdens{\D A}} + \vals{\incdens{\D B}}
      = \vals{\incden{\D A + \D B}}
      = \vals{\incdens{\D(A + B)}}
      \).

    \item[Case $A \to B$:]
      \begin{align*}\D\incden{A \to B}
        &= \D(\expO{\incdens A}{\incdens B})\\
        &= \expO{\isof{\vals{\incdens
              A}}}{\expO{\D\incdens{A}}{\D\incdens{B}}}\\
        &= ...\\
        &= \CP(\isof{\incdens A},\ {\expO{\incdens{\D A}}{\incdens{\D B}}})\\
        &= \vals{(\expO{\isof{\incdens{A}}}{\expO{\incdens{\D A}}{\incdens{\D B}}})}\\
        &= \vals{\incden{\isof A \to \D A \to \D B}}\\
        &= \vals{\incdens{\D(A \to B)}}
      \end{align*}

    \item[Case $\tunit$:]
  \end{description}
\end{proof}


\section{Semi\naive{} evaluation}
\label{sec:seminaive}

\todo{TODO: explain \prim{fast-fix}. explain $\Phi$ type transformation as
  ``decorating'' $\isof A$ values with their zero changes so that \prim{fix}
  gets its argument's derivative and can be implemented via \prim{fast-fix}.}

\Cref{fig:seminaive-phi,fig:seminaive-delta} define two mutually recursive
static transformations: $\phi e$, which computes $e$ \emph{semi\naive{}ly},
speeding up fixed point computation by using derivatives; and $\delta e$, which
computes the \emph{derivative} of $\phi e$.
%
To describe the types of $\phi e$ and $\delta e$, we'll need three new operators
on contexts: $\isocx\G$, $\D\G$, and $\Phi\G$.
%
We define these by their pointwise action on hypotheses:

\nopagebreak[2]
\begin{align*}
  \isocx{(\h x A)} &= \hd x A & \isocx{(\hd x A)} &= \hd x A\\
  \D(\h x A) &= \h \dx {\D A} & \D(\hd x A) &= \emptycx\quad\text{(the empty context)}\\
  \Phi(\h x A) &= \h x {\Phi A} & \Phi(\hd x A) &= \hd x {\Phi A}, \hd \dx {\DP A}
\end{align*}

\noindent For example,
%
\begin{mathpar}
  \isocx{(\h x A, \hd y B)} = \hd x A, \hd y B

  \D(\h x A, \hd y B) = \h \dx {\D A}

  \Phi(\h x A, \hd y B) = \h x {\Phi A}, \hd y {\Phi B}, \hd \dy {\DP B}
\end{mathpar}

\noindent
The idea is that $\isof\G$, $\D\G$, and $\Phi\G$ mirror the effect of $\iso$,
$\D$, and $\Phi$ on the semantics of $\G$. In particular:

\nopagebreak[2]
\begin{align*}
  \den{\isocx\G} &\cong \isof{\den\G}
  &
  \den{\Phi(\h x A)} &\cong \den{\Phi A}
  \\
  \den{\D\G} &\cong \D\incdens\G
  &
  \den{\Phi(\hd x A)} &\cong \den{\Phi \isof A}
\end{align*}

\noindent
These defined, we can state the sense in which $\phi$ and $\delta$ are
type-correct:

\begin{restatable}[Type-correctness]{theorem}{thmTypeCorrect}
  \label{thm:type-correct}
  If $\J e \G A$, then
  \[ \J {\phi e} {\Phi\G} {\Phi A}
  \qquad\text{and}\qquad
  \J {\delta e} {\isocx{\Phi\G}, \DP\G} {\DP A}
  \]
\end{restatable}
\begin{proof}
  By induction on typing derivations; see \cpageref{proof:typeCorrect}.
\end{proof}

\noindent To get the hang of these context and type transformations, let's
suppose that $\J e {\h x A, \hd y B} C$. Then \cref{thm:type-correct} tells us:

\nopagebreak[2]
\begin{align*}
  \Jalign {\phi e} {\h x {\Phi A}, \hd y {\Phi B}, \hd \dy {\DP B}} {\Phi C}
  \\
  \Jalign {\delta e} {\hd x {\Phi A}, \h\dx{\DP A}, \hd y {\Phi B}, \hd \dy {\DP B}} {\DP C}
\end{align*}


\input{seminaive/fig-seminaive-phi}
\input{seminaive/fig-seminaive-delta}


\section{Logical relation}

\begin{definition}\label{def:approx}\label{def:approx-contexts}
  We define a type-indexed logical relation ${\impls_A} \subseteq
  \vals{\incdens{\Phi A}} \x \vals{\incdens{A}}$ as follows:

  \nopagebreak[2]
  \begin{align*}
    %% unit & products
    \tuple{x_i}_i \impls \tuple{y_i}_i &\iff \fa{i} x_i \impls y_i
    \\
    %% sums
    \inj i x \impls \inj j y &\iff i = j \wedge x \impls y
    \\
    %% functions
    f \impls_{A \to B} g &\iff \fa{x \impls_A y} f\<x \impls_B g\<y
    \\
    %% boxes
    \tuple{x,\dx} \impls_{\isof A} y
    &\iff x \impls_A y \wedge \vld{\dx} x x
    \\
    %% sets
    s \impls_{\tseteq A} t
    %% &\iff (\forall x \in s, \exists y \in t)~x\approx_{\eqt A} y
    %% \wedge (\forall y \in t, \exists x \in s)~x \approx_{\eqt A} y
    &\iff \faex{x \in s}{y \in t} x\approx_{\eqt A} y
    \wedge \faex{y \in t}{x \in s} x \approx_{\eqt A} y
  \end{align*}

  \noindent
  This lifts to contexts, ${\impls_\G} \subseteq \den{\Phi\G} \x \den\G$, as
  follows:

  \nopagebreak[2]
  \begin{align*}
    \gamma \impls_\G \rho
    &\iff \fa{\h x A \in \G} \gamma_x \impls \rho_x
    \\ &\phantom{\iff}
    \wedge \fa{\hd x A \in \G}
    (\gamma_{\dvar x} \impls \rho_{\dvar x}
    \wedge \vld{\gamma_{\dvar{\dx}}}{\gamma_{\dvar x}}{\gamma_{\dvar x}})
  \end{align*}
\end{definition}

%% \noindent
%% Which can be lifted to morphisms (and thus term denotations) as follows:

%% \begin{align*}
%%   \impls_{\G \vdash A} &\subseteq \den{\Phi\G \vdash \Phi A} \x \den{\G \vdash A}\\
%%   f \impls_{\G \vdash A} g &\iff \fa{\gamma \impls_\G \gamma'} f\<\gamma \impls_A g\<\gamma'
%% \end{align*}

\begin{restatable}[Correctness]{theorem}{thmCorrect}\label{thm:correct}
  If $\J e \G A$ and $\gamma \impls_\G \rho$ then

  \begin{enumerate}
  \item \(\den{\phi e} \<\gamma \impls_A \den{e} \<\rho\).
  \item If $\vld{\dgamma}{\g}{\g'}$ then \(\vld{\den{\delta e}
    \tuple{\gamma,\dgamma}}{\den{\phi e} \<\gamma}{\den{\phi e} \<\gamma'}\).
  \end{enumerate}

  %% \nopagebreak[2]
  %% \begin{gather*}
  %% \text{(1)}~~ \den{\phi e}\<\gamma \impls_A \den{e}\<\rho
  %% \qquad\text{and}\qquad
  %% \text{(2)}~~
  %% \vld{\den{\delta e} \tuple{\gamma,\dgamma}}{\den{\phi e} \<\gamma}{\den{\phi e} \<\gamma'}
  %% : \incdens{\Phi{A}}
  %% \end{gather*}

  \noindent
  Note that $\tuple{\gamma,\dgamma}$ is a slight abuse of notation here;
  \todo{TODO explain that \emph{monotone} variables in $\g$ when we use it as
    $\den{\phi e}\<\g$ turn into \emph{discrete} variables in $\extend\g\dgamma$
    when we use it in $\den{\delta e} \<\extend\g\dgamma$}.
\end{restatable}
\begin{proof}
  By induction on the derivation of $\J e \G A$. See \cpageref{proof:correct}.
\end{proof}


