\section{Related and Future Work}
\label{sec:futurework}

\paragraph{Aggregation}
Aggregation of values --- for example, taking the sum $\sum_{x \in A} f \;x$ of
a function $f$ across a set $A$ --- is a useful and ubiquitous database
operation. Datafun naturally supports \emph{semilattice} aggregation via
$\bigvee$, but many natural operations such as summation do not form
semilattices on their underlying type. There are several potential ways to add
support for aggregations to Datafun:
\begin{itemize}
\item Common aggregations can be provided as primitive functions, for example
  $\ms{size} : \Set{\eq{A}} \mto \N$ or $\ms{sum} : (\eq{A} \to \N) \uto
  \Set{\eq{A}} \mto \N$.

\item In the style of Machiavelli~\cite{machiavelli}, one could add a general
  operator $\ms{hom} : B \to (A \uto B \uto B) \uto \Set{A} \uto B$, which
  effectively linearizes a set in an unspecified order. The semantics of
  \ms{hom} are, alas, necessarily nondeterministic.

\item One could augment Datafun with a type of bags (multisets) $\Bag{A}$; bags
  naturally support a much broader class of aggregation --- commutative monoids
  --- than sets. See, for example, \citet{multilinear-bigdata} and
  \citet{reladj}.
\end{itemize}

\paragraph{Optimization} Because Datalog is so strongly constrained,
there has been a lot of very successful work on optimizing it, ranging
from compilation into binary decision diagrams~\cite{bdd} by
\citet{whaley-lam}, to the famous ``magic sets''~\cite{magicsets}
algorithm.

From our perspective, magic sets are a natural next step for
investigation into how to optimize Datafun. Intuitively, the magic
sets algorithm exploits the fact that Datalog (as a total logic
language) has both a top-down and bottom-up reading, and rewrites the
program so that it does bottom-up search, while using top-down
reasoning to strategically avoid adding useless facts to the
database. Transplanting this analysis to Datafun would essentially
give us optimized implementations of fixed points, but extending the
magic sets algorithm is likely to be very subtle, since Datafun has
higher-order functions and Datalog does not. As a result, our goal is
to first see if magic sets can be applied to first-order Datafun programs,
and then use defunctionalization~\cite{defunctionalization} to
extend it to full Datafun.

Very recently, \citet{flix} have introduced the Flix language, which
extends the semantics of Datalog to support defining relations valued
in arbitrary lattices (rather than just the powerset of atoms). Like
Datafun, this lets Flix support using monotone functions (on suitable
lattices) in program expressions. Unlike Datafun, Flix does not yet
have monotonicity checking for programmer-defined operators. However,
because Flix does not extend Datalog to higher order, efficient
Datalog implementation strategies (such as semi-naive evaluation)
continue to apply.


\paragraph{Databases} Datalog has sometimes been described
as ``relational algebra plus fixed points'', and there is a long line
of work on embedding database query languages into general-purpose
languages, including pioneering efforts such as
Machiavelli~\cite{machiavelli} and Kleisli~\cite{kleisli}, as well as
more recent systems such as Ferry~\cite{ferry} and LINQ in C\#~\cite{linq-wadler}.
%
The focus of this work has been on embedding query languages based
on relational algebra into general purpose languages, with an emphasis
on statically compiling higher-order queries into the first-order
queries supported by existing database systems (\citet{query-shredding} is a
representative example).

Our approach is a little bit different. Instead of embedding Datalog
into a general purpose language, Datafun is \emph{also} a ``little
language'', albeit one that happens to be a higher-order
functional language. We very deliberately did not try to embed Datafun
into an existing language, because that would have greatly complicated
the context-management operations needed to ensure monotonicity.

In fact, from a language designer's perspective, Datafun can be seen
as an argument in favor of extending functional languages to support
programming with user-defined, non-strong comonads.

\paragraph{Deletion} \citet{logical-algorithms} showed how
forward-chaining logic programming permits concise and elegant
expression of a wide variety of algorithms, including a natural cost
semantics. However, they noted that there were some algorithms (such
as union-find and greedy algorithms) which could be formulated in this
style, \emph{if} there were additionally support for deleting facts
from a database. Later, \citet{linear-logical-algorithms} went on to
show how deletion could be given a logical interpretation by
formulating in terms of linear logic programming.

This naturally raises the question of whether we could identify a
``linear Datafun'' corresponding to this style of programming, where
we might linear types to model features like deletion. There are many
nontrivial semantic issues (e.g., how to define monotonicity), but
it seems a promising question for future work.

\paragraph{Termination}

Datafun as presented is Turing-incomplete. This is advantageous for
optimization; for example, one powerful optimization technique is \emph{loop
  reordering} (in SQL terminology, \emph{join reordering}), that is, taking
advantage of the equation
\begin{eqnarray*}
  \forin{x \in e_1} \forin{y \in e_2} e
  &=& \forin{y \in e_2} \forin{x \in e_1} e
\end{eqnarray*}
when $x,y \notin \ms{FV}(e_1) \cup \ms{FV}(e_2)$. But this equation does not
always hold in the presence of nontermination; for example, if $e_1 = \unit$ and
$e_2$ diverges.

Nonetheless, without adding advanced facilities for termination
checking, there are many functions it is difficult to implement
without use of general recursion. So a natural direction for future
work is to study how to add support for general recursion to Datafun.
Because domains~\cite{domain-theory} can be understood as partial
orders with directed joins, there are likely many interesting
categorical structures connecting the category of domains to the
category of posets, some of which will hopefully lead to a principled
type-theoretic integration of partial functions into Datafun.

\paragraph{User-Defined Posets and Semilattices}
The two fundamental semilattice types Datafun provides are booleans and sets;
products and functions merely preserve semilattice structure where they find it.
One might contemplate allowing the programmer to define their own semilattice
structures using something like Haskell's \texttt{newtype}/\texttt{instance}. In
general, this is a difficult problem, because we may need to do serious
mathematical reasoning to prove that a comparison function implements a partial
ordering, or that a datatype can be equipped with a semilattice structure
obeying this partial ordering which is commutative, associative and idempotent.

One example of such a family of types are the \emph{lexicographic sum
  types}. Given two posets $P$ and $Q$, their disjoint union $P + Q$
is also a poset, with left values compared by the $P$-ordering, and
right values compared by the $Q$-ordering, and no ordering between
left and right values. However, this is not the only way that the
disjoint union could be equipped with an order structure.

For example, we could define the \emph{lexicographic} sum $P \lsum Q$,
which has the same elements as the sum, but extending the coproduct order
relation with the additional facts that $\ms{in}_1(p) \leq \ms{in}_2(q)$.
Indeed, we already have a special case of this: as we noted earlier, our boolean
type is not $1 + 1$, but it \emph{is} $1 \lsum 1$.

But as our Booleans already show, giving good syntax for their
eliminators is difficult, because we have to show that not just a term
is monotone, but that the different branches of a lexicographic case
expression are ordered with respect to \emph{each other}. For the case
of ordered Booleans, we were able to give a special eliminator which
guaranteed it, but in general it requires proof.

One natural direction for future work is to extend the syntax of
Datafun with support for these kinds of proofs, perhaps taking
inspiration from dependent type theory.
