%% FIGURE: Typing rules
\begin{figure*}
  %% \boxed{\ensuremath{\mathsz{10pt}{\J{\Delta}{\Gamma}{e}{A}}}}
  \begin{mathpar}
    \infer[\rn{\rt{var}}]{\J{\GD}{\GG}{x}{A}}{x\of A \in \GD}
    \and
    \infer[\rn{\rt{var}^+}]{\J{\GD}{\GG}{\m{x}}{A}}{\m{x}\of A \in \GG}
    %% function rules
    \and
    \infer[\rn{\fn}]{\J{\GD}{\GG}{\fn\bind{x} e}{A \uto B}}{
      \J{\GD,x\of A}{\GG}{e}{B}}
    \and
    \infer[\rn{\rt{app}}]{\J{\GD}{\GG}{e_1\;e_2}{B}}{
      \J{\GD}{\GG}{e_1}{A \uto B} &
      \J{\GD}{\cdot}{e_2}{A}}
    \and
    \infer[\rn{\monofn}]{\J{\GD}{\GG}{\fn\bind{x} e}{A \mto B}}{
      \J{\GD}{\GG,x \of A}{e}{B}}
    \and
    \infer[\rn{\rt{app}^+}]{\J{\GD}{\GG}{e_1\;e_2}{B}}{
      \J{\GD}{\GG}{e_1}{A \mto B} &
      \J{\GD}{\GG}{e_2}{A}}
    %% product & sum rules
    \and
    \infer[\rn{\rt{pair}}]{\J{\GD}{\GG}{(e_1,e_2)}{A_1 \x A_2}}{
      \J{\GD}{\GG}{e_i}{A_i}}
    \and
    \infer[\rn{\pi}]{\J{\GD}{\GG}{\pi_i\;e}{A_i}}{\J{\GD}{\GG}{e}{A_1 \x A_2}}
    \and
    \infer[\rn{\rt{in}}]{\J{\GD}{\GG}{\ms{in}_i\;e}{A_1 + A_2}}{
      \J{\GD}{\GG}{e}{A_i}
    }
    \and
    \infer[\rn{\rt{case}}]{\J{\GD}{\GG}{\case{e}{x}{e_1}{x}{e_2}}{C}}{
      \J{\GD}{\cdot}{e}{A_1 + A_2} &
      \J{\GD,x\of A_i}{\GG}{e_i}{C}}
    \and
    \infer[\rn{\rt{case}^+}]{
      \J{\GD}{\GG}{\case{e}{\m{x}}{e_1}{\m{x}}{e_2}}{C}
    }{
      \J{\GD}{\GG}{e}{A_1 + A_2} &
      \J{\GD}{\GG,\m{x}\of A_i}{e_i}{C}
    }
    %% boolean rules
    %% \and
    %% \infer[\rn{=}]{\J{\GD}{\GG}{e_1 = e_2}{\bool}}{\J{\GD}{\GG}{e_i}{\eq{A}}}
    \and
    \infer[\rn{\rt{true}}]{\J{\GD}{\GG}{\ms{true}}{\bool}}{}
    \and
    \infer[\rn{\rt{false}}]{\J{\GD}{\GG}{\ms{false}}{\bool}}{}
    \and
    \infer[\rn{\rt{if}}]{\J{\GD}{\GG}{\ifthen{e}{e_1}{e_2}}{A}}{
      \J{\GD}{\cdot}{e}{\bool} &
      \J{\GD}{\GG}{e_i}{A}}
    \and
    \infer[\rn{\rt{if}^+}]{\J{\GD}{\GG}{\ifthen{e}{e_1}{\unit}}{L}}{
      \J{\GD}{\GG}{e}{\bool} &
      \J{\GD}{\GG}{e_1}{L}}
    %% set & semilattice rules
    \and
    \infer[\unit]{\J{\GD}{\GG}{\unit}{L}}{}
    \and
    \infer[\rn{\vee}]{\J{\GD}{\GG}{e_1 \vee e_2}{L}}{\J{\GD}{\GG}{e_i}{L}}
    \and
    \infer[\rn{\{\}}]{\J{\GD}{\GG}{\{e\}}{\Set{A}}}{\J{\GD}{\cdot}{e}{A}}
    \and
    \infer[\rn{\bigvee}]{\J{\GD}{\GG}{\letin{x}{e_1}{e_2}}{L}}{
      \J{\GD}{\GG}{e_1}{\Set{A}} &
      \J{\GD,x\of A}{\GG}{e_2}{L}}
    %% \and
    %% \infer[\rn{\{:\}}]{\J{\GD}{\GG}{\singlemap{e_1}{e_2}}{\Map{A}{B}}}{
    %%   \J{\GD}{\cdot}{e_1}{A} &
    %%   \J{\GD}{\GG}{e_2}{B}}
    \and
    \infer[\rn{\rt{fix}}]{\J{\GD}{\GG}{\fix{\m{x}}{e}}{\fineq{L}}}{
      \J{\GD}{\GG,x\of L}{e}{\fineq{L}}}
    \and
    \infer[\rn{\rt{fix}_{\le}}]{
      \J{\GD}{\GG}{\fixle{\m{x}}{e_1}{e_2}}{\eq{L}}
    }{
      \J{\GD}{\GG}{e_1}{\eq{L}} &
      \J{\GD}{\GG,x \of \eq{L}}{e_2}{\eq{L}}}
  \end{mathpar}

  \caption{Typing rules for core Datafun}
  \label{fig:typing-rules}
\end{figure*}


\section{Typing rules}

Datafun's typing judgment $\J{\GD}{\GG}{e}{A}$ is defined by the inference rules
given in Figure \ref{fig:typing-rules}. We gloss $\J{\GD}{\GG}{e}{A}$ as
follows: ``expression $e$ has type $A$ using variables from $\GD \cup \GG$, and
moreover the value of $e$ is \emph{monotone} with respect to the variables in
$\GG$''.

The context $\GD$ types ordinary variables; $\GG$, monotone variables. Both
admit the usual structural rules of exchange, weakening, and contraction.
Variables from either context may be used freely (rules \rt{var}, $\rt{var}^+$).
\todo{explain?}

\subsection{Functions and application}
Two function types require two function introduction rules: the ordinary
$\lambda$ and the monotone $\lambda^+$. These simply introduce variables into
their respective contexts. Monotone function application $\rt{app}^+$ is
perfectly standard, but ordinary function application \rt{app} has a pecularity:
the argument $e_2$ gets an \emph{empty} monotone context.

To understand why, recall our gloss: the application $e_1\;e_2$ must be monotone
in $\GG$. But $e_1$ is an ordinary, and in general \emph{non-monotone}, function
$A \uto B$: there is no guarantee that it respect any order on its argument.
(Suppose, for example, $e_2$ were some monotone variable $\m{x} : A \in \GG$.)
We work around this scoff-law behavior on $e_1$'s part by ensuring its argument
$e_2$ is \emph{constant} with respect to $\GG$---which we accomplish by simply
prohibiting $e_2$ from using any of $\GG$'s variables.

This technique of \emph{wiping clean} the monotone context to guarantee
constancy\footnote{Wherever we write ``constant'' in this section, substitute
  ``constant with respect to the monotone context''. The ordinary context is
  never ``wiped clean'', and behaves entirely as it would in a simply-typed
  $\lambda$-calculus.} of a subterm recurs in several other rules. Readers
familiar with judgmental formulations of modal logics of necessity such as
\citet{jrml} may notice a feeling of \textit{d\'ej\`a vu}; indeed, there is a
hidden comonad at work here. But we are getting ahead of ourselves. For more on
that, turn to Section \ref{sec:semantics}.

\todo{Get Neel to provide more citations here?}

\todo{Mention comonads and modal logic?}

\subsection{Products and sums}
The pairing and projection rules, \rt{pair} and $\pi$, are completely standard,
as is the \rt{in} rule for sum introduction. Sum elimination, however, splits
into two rules, \rt{case} and $\rt{case}^+$. $\rt{case}^+$ requires its branches
to be monotone in the variable $\m{x}$ it introduces, and consequently its
subject $e$ is permitted access to the monotone context $\GG$. \rt{case},
however, analyses its subject $e$ as a constant --- wiping clean its monotone
context --- and thus is allowed to introduce the variable $x$ into the ordinary
context $\GD$. \todo{rewrite for clarity}

\subsection{Booleans}
While \rt{true} and \rt{false} are straightforward, there are two rules for
boolean elimination, \rt{if} and $\rt{if}^+$. This is because in Datafun, $1$
plus $1$ does not equal $\bool$: booleans are \emph{not} a sum of
units.\footnote{For simplicity, we have omitted the unit type 1 from our
  presentation of Datafun, but it is easy enough to imagine including it.} At
the type $1 + 1$, $\ms{in}_1 \triv$ and $\ms{in}_2 \triv$ are incomparable. But
in Datafun, $\ms{true} > \ms{false}$. Therefore, to eliminate a boolean in a
monotone fashion, one must ensure one's \emph{then}-branch is always greater
than one's \emph{else}-branch.

Thus Datafun has two \ms{if} rules. First, \rt{if}, where the boolean $e$ being
analysed is constant (has an empty monotone context), and so the branches $e_1$,
$e_2$ may be arbitrary expressions.

Second, $\rt{if}^+$, where the subject $e$ has full access to $\GG$, but the
\ms{if}-expression must have \emph{semilattice type}, and the \emph{else}-branch
is constrained to be $\unit$ --- the least value, thus smaller than $e_1$.

This is a conservative approach: there are many semantically monotone, but
untypeable, \ms{if}-terms. However, it is complete for semilattice types, for in
that case $(\ifthen{e}{e_1}{e_2})$ may be rewritten $(e_2 \vee
\ifthen{e}{e_1}{\unit})$; as long as $e_1 \ge e_2$ and so $e_2 \vee e_1 = e_1$,
this will not change the meaning of the expression, only (potentially) its
execution efficiency.

Thus the only meaningful restriction here is to semilattice types. In practice,
we have yet to find a case where this is problematic.

\todo{Mention that $e_1 \ge e_2$ is in general statically undecidable (rice's
  theorem), and that a typesystem to check it would amount to dependent typing?}

\subsection{Semilattices and sets}
The semilattice $\unit$ and $\vee$ operations are typed by their
correspondingly-named rules. As $\vee$ is monotone, its arguments have full
access to the monotone context $\GG$.

Recall that sets are ordered by inclusion: although $2 \le 3$, nonetheless
$\singleset{2} \not\le \singleset{3}$. For this reason the rule $\{\}$ for
constructing a singleton set $\singleset{e}$ wipes clean its element $e$'s
monotone context. Datafun does not need empty-set or union operators, since
$\unit$ and $\vee$ generalize them.

Finally, we come to $\bigvee$, the set-comprehension rule. This rule has the
flavor of a monadic ``bind'' operation, but generalized to a result of any
semilattice type. This operation is naturally monotone both in the set $e_1$
being iterated over and in the expression $e_2$ which we are taking the least
upper bound of. Moreover, $e_2$ is \emph{not} required to be monotone in the
elements $x \in e_1$. \todo{Why not?}

%% \subsection{Maps}
%% Finite maps are an abstract type, and most of their interface is exposed through
%% primitive functions (see Figure \ref{fig:primitives}). The only typing rule is
%% $\{:\}$, the introduction rule for singleton maps. By analogy with the
%% introduction rule for singleton sets $\{\}$,

\subsection{Fixed points}

$\rt{fix}_\le$ exploits the fact that a

\newpage
